---
title: "Average distance"
author: "Josef Klafka and Daniel Yurovsky"
date: "3/24/2020"
output: html_document
---

In this script, we quantify the average pairwise distance between unigram and trigram barycenters. This simple summary statistic (computed along with a non-parametric bootstrap) will give evidence that the unigram barycenters have more variation than the trigram barycenters. 

```{r setup, include=FALSE}
require(tidyverse)
require(tidyboot)
require(here)
require(reshape2)

knitr::opts_chunk$set(echo = TRUE)
```

We read in the unigram and trigram barycenters. 

```{r read in data}
barycenters <- read_csv(here("Data/5barycenters.csv")) %>% 
  select(-language)

unigrams <- barycenters %>%
  filter(source == "wikipedia", gram == "unigram")

trigrams <- barycenters %>%
  filter(source == "wikipedia", gram == "trigram")
```

We use the dynamic time warping distance metric to quantify the differences between the barycenters. The average unigram distance is $3.04$, with a upper confidence interval bound of $3.08$ and a lower confidence interval bound of $3.01$. Those values are $2.50$, $2.52$ and $2.49$ respectively for trigrams. Qualitatively, there appears to be a large difference in means. 

```{r get distances}
dist_measure <- "DTW"

unigram_dist <- unigrams %>% 
  select(-source, -gram) %>%
  proxy::dist(method = dist_measure) %>%
  as.matrix() %>%
  melt(varnames = c("row", "col")) %>%
  as_data_frame() %>%
  filter(row > col) %>%
  select(-row, -col)

trigram_dist <- trigrams %>% 
  select(-source, -gram) %>%
  proxy::dist(method = dist_measure) %>%
  as.matrix() %>%
  melt(varnames = c("row", "col")) %>%
  as_data_frame() %>%
  filter(row > col) %>%
  select(-row, -col)

dists <- unigram_dist %>%
  mutate(gram = "unigram") %>%
  bind_rows(trigram_dist %>% mutate(gram = "trigram"))

## get nonparametric bootstrap over unigram and trigram distances
boots <- unigram_dist %>% 
  tidyboot_mean(value) %>%
  mutate(gram = "unigram") %>%
  bind_rows(trigram_dist %>% tidyboot_mean(value) %>% mutate(gram = "trigram"))
```

We run a one-sided t-test to see if the mean of the unigram distances is greater than the mean of the trigram distances to a statistically significant level. The p-value is $4.386501 x 10^{-157}$. 

```{r run t-test for differences in distances}
t.test(unigram_dist %>% pull(value), trigram_dist %>% pull(value), 
       alternative = "greater", var.equal = F)$p.value
```

A violin and boxplot to display the distribution of distances. The violin plots for unigrams and trigrams actually appear similarly distributed, but unigrams has more large distance values (a fatter tail, in probability distribution terms) which allocates mass away from small distance values that dominate both trigrams and unigrams. 

```{r plot}
dists %>%
  ggplot(aes(x = gram, y = value)) + 
    geom_violin(aes(fill = gram)) + 
    geom_boxplot(width = 0.1) + 
    coord_flip() +
    ggtitle("Distribution of pairwise DTW distances between Wikipedia barycenters")
```



