---
title: "CHILDES smoothed analysis"
author: "Josef Klafka and Daniel Yurovsky"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
require(tidyverse)
require(here)
require(feather)
require(childesr)

knitr::opts_chunk$set(echo = TRUE)
```

```{r childes unigrams}
eng <- get_utterances(collection = "Eng-NA")
```

```{r plotting trigrams data}
childes <- read_feather(here("Data/CHILDES/childes_results.feather"))

childes %>%
  # filter(language == "Eng-NA") %>%
  filter(length == 7) %>% 
  ggplot(aes(x = position, y = surprisal, color = speaker)) + 
    geom_point(show.legend = F) + 
    geom_line() + 
    facet_wrap(~language)

# scale to percentage of distribution
childes %>%
  group_by(language, speaker, length) %>% 
  mutate(percent_s = surprisal / sum(surprisal)) %>%
  ungroup() %>%
  filter(length == 8, speaker == "Adult") %>%
  ggplot(aes(x = position, y = percent_s, group = language, color = language)) + 
    geom_point(show.legend = F) + 
    geom_line() + 
    facet_wrap(~language)
```

Analyzing British National Corpus (BNC) and Switchboard results. 
```{r bnc and switch}
switch_results <- read_csv(here("Data/corpus/Switchboard_results.csv"), 
                           col_names = c("position", "surprisal", "length"))
bnc_results <- read_csv(here("Data/corpus/bnc_results.csv"), 
                           col_names = c("position", "surprisal", "length"))

bnc_results %>%
  filter(length %in% 20:25) %>% 
  ggplot(aes(x = position, y = surprisal)) + 
    geom_point(show.legend = F) + 
    geom_line() + 
    facet_wrap(~length)
```


A little over 30 percent of the utterances in Switchboard are only one or two words (probably "yeah" or "um"). There's a long tail after that, almost completely flattening out by 40 words. 
```{r sentence length distribution for switchboard}
bnc <- read_table(here("Data/corpus/trimmed_bnc.txt"), col_names = "text")
switchboard <- read_table(here("Data/corpus/Switchboard.txt"), col_names = "text")

switchboard %>% 
  mutate(length = str_count(text, boundary("word"))) %>%
  count(length) %>% 
  mutate(n = n/sum(n)) %>%
  ggplot(aes(x = length, y = n)) + 
    geom_bar(stat = "identity")
```

The below code gets together the results from the CHILDES pipeline (with all languages given in the `langs` vector), labels the language and speaker role and writes everything into a single feather. You can change the input languages based on what you've run and the output name for clarity. 
```{r knitting together childes results}
# files <- list.files(path = "Data/CHILDES/",pattern = ".csv")
# map_dfr(files, ~read_csv(here(paste0("Data/CHILDES/", .)), 
#                          col_names = c("position", "surprisal", "length")))
langs <- c("Chinese", "French", "German", "Spanish", "Japanese", "Eng-NA")
output_name <- c("childes_results")

get_childes_results <- function(lang) {
  read_csv(here(paste0("Data/CHILDES/", lang, "_adult_results.csv")),
                    col_names = c("position", "surprisal", "length")) %>%
    mutate(speaker = "Adult") %>%
    bind_rows(
      read_csv(here(paste0("Data/CHILDES/", lang, "_child_results.csv")),
                      col_names = c("position", "surprisal", "length")) %>%
        mutate(speaker = "Child")
    ) %>%
    mutate(language = lang)
}

childes_results <- map_dfr(langs, ~get_childes_results(.))

childes_results %>% 
  write_feather(here(paste0("Data/CHILDES/", output_name, ".feather")))
```